Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-09-09 paper/main.tex
example derivation

% Comment for anonymization
% Add preprint option for the extended version
% Also look for "SHORT" comments
% \def\fullversion{}

@@

>>>>

% Comment for anonymization
% Add preprint option for the extended version
% Also look for "SHORT" comments
\def\fullversion{}

@@

\section{Example Derivation in the Automatic System}
\label{app:derivation}

\newcommand{\loopbody}[0]{\ensuremath{\code{if~(x - 10 \geq 0)~\{x=x{-}10; tick(5);\}~else~break;}}}

\begin{figure*}[t]
  \centering
  \footnotesize
  \begin{mathpar}
  \inferrule*[right={\scriptsize (Q:Loop)},leftskip=0cm,rightskip=0cm,width=20cm]
  { \inferrule*[right={\scriptsize (Q:If)},leftskip=0cm,rightskip=0cm,width=20cm]
    { 
      \inferrule*[right={\scriptsize (Q:Weak)},leftskip=2cm,rightskip=4cm, vdots=1.5cm]
      { 
        \inferrule*[right={\scriptsize (Q:Seq)},leftskip=0cm,rightskip=0cm]
        {
\inferrule*[right={\scriptsize (Q:Weak)},leftskip=0cm,rightskip=0cm]
{
  \inferrule*[right={\scriptsize (Q:DecP)},leftskip=0cm,rightskip=0cm]
  { }
  {(x {<} 10, B^\text{de}); ((x {\geq} 10,Q^\text{de})  \vdash \code{x = x - 10}
   \dashv (\cdot, P^\text{de})
  }
}
{(x {<} 10, B^\text{we}); ((x {\geq} 10,Q^\text{we})  \vdash \code{x = x - 10}
\dashv (\cdot, P^\text{we})
}
\and
\inferrule*[right={\scriptsize (Q:Tick)},leftskip=0cm,rightskip=0cm]
{ }
{(x {<} 10, B^\text{ti}); (\cdot,Q^\text{ti})  \vdash \code{tick(5)}
 \dashv (\cdot, P^\text{ti})
}
} {
          (x {<} 10, B^\text{sq}); ((x {\geq} 10,Q^\text{sq}) \vdash \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{sq}) 
        }
      }
      { (x {<} 10, B^\text{if}); ((x {\geq} 10,Q^\text{if}) \vdash \code{x=x{-}10; tick(5);} \dashv (\cdot,P^\text{if}) }
      \and
      \inferrule*[right={\scriptsize (Q:Weak)},leftskip=4.5cm,rightskip=4cm]
      { \inferrule*[right={\scriptsize (Q:Break)},leftskip=0cm,rightskip=0cm]
        { }
        {(x {<} 10, B^\text{br}); (x {<} 10,Q^\text{br}) \vdash \code{break} \dashv (\bot,P^\text{br})}
      }
      {
        (x {<} 10, B^\text{el}); (x {<} 10,Q^\text{el}) \vdash \code{break} \dashv (\cdot,P^\text{el})
      }
    }
    { \hspace{3cm} (x {<} 10, B^\text{lo}); (\cdot,Q^\text{lo}) \vdash \loopbody \dashv (\cdot,P^\text{lo}) \hspace{3cm} 
    }
  }
  {
   (\cdot, B); (\cdot, Q) \vdash \code{loop}~{\loopbody} \dashv (x {<} 10, P) 
  }
\end{mathpar}
\flushleft Constraints:
$$
\begin{array}{llll}
P = B^\text{lo} & Q  = Q^\text{lo} & Q^\text{lo} = P^\text{lo} \\
B^\text{el} = B^\text{if} = B^\text{lo} & Q^\text{el} = Q^\text{if} = Q^\text{lo} & P^\text{el} = P^\text{if} = P^\text{lo} \\
B^\text{el} = B^\text{br}  & Q^\text{el} \succeq_{(x{<}10)} Q^\text{br}  & P^\text{br} \succeq_{(\cdot)} P^\text{el} \\
B^\text{br} = Q^\text{br}\\
B^\text{if} = B^\text{sq}  & Q^\text{if} \succeq_{(x{<}10)} Q^\text{sq}  & P^\text{sq} \succeq_{(\cdot)} P^\text{if} \\
B^\text{sq} = B^\text{we} = B^\text{ti} & Q^\text{sq} = Q^\text{we} & P^\text{we} = Q^\text{ti} & P^\text{ti} = P^\text{sq} \\
Q^\text{ti} = P^\text{ti} + 5 \\
B^\text{we} = B^\text{de}  & Q^\text{we} \succeq_{(x{<}10)} Q^\text{de}  & P^\text{de} \succeq_{(\cdot)} P^\text{we} \\
p^\text{de}_{0,10} = q^\text{de}_{0,10} + q^\text{de}_{0,x}  & p^\text{de}_{0} = q^\text{de}_{0}
& \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{de}_{\alpha,\beta} = q^\text{de}_{\alpha,\beta}
\end{array}
$$
\flushleft Linear Objective Function:
$$
q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x}
$$
\flushleft Constant Objective Function:
$$
q_0 + 11{\cdot}q_{0,10}  
$$
\caption{An example derivation as produced by the automatic tool.}
\label{fig:derivation}
\end{figure*}

\pref{fig:derivation} contains an example derivation as it is produced
by our automatic analyzer.  The upper case letters (with optional
superscript) such as $Q^\text{de}$ stand for a family of rational
variables that are later part of the constraint system that is passed
to the LP solver.  For example $Q^\text{de}$ stands for the potential
function $q^\text{de}_0 + q^\text{de}_{x,0} |[x,0]| +
q^\text{de}_{0,x} |[0,x]| + q^\text{de}_{x,10} |[x,10]| +
q^\text{de}_{10,x} |[10,x]| + q^\text{de}_{0,10} |[0,10]|$, where the
variables such as $q^\text{de}_{x,10}$ are yet unknown and later
instantiated by the LP solver.  The derivation with the instantiated
variables is then a sound derivation in the verified program logic.

The rules are syntax directed and applied inductively.  That means
that we follow the inductive definition of the syntax.  For example,
the outermost expression is a loop, so we use the rule \text{Q:Loop}
at the root of the derivation tree.  At this point, we don't know yet
whether a loop invariant exists.  But we produce the constraints
$Q^\text{lo} = P^\text{lo}$ which is short for the following
constraint set.
$$
\begin{array}{lll}
q^\text{lo}_0 = p^\text{lo}_0 &
q^\text{lo}_{x,0}= p^\text{lo}_{x,0} &
q^\text{lo}_{0,x}= p^\text{lo}_{0,x} \\
q^\text{lo}_{x,10}= p^\text{lo}_{x,10} &
q^\text{lo}_{10,x}= p^\text{lo}_{10,x} &
q^\text{lo}_{0,10}= p^\text{lo}_{0,10}
\end{array}
$$
%
This constraint set expresses the fact that the potential before and
after the loop body is equal and thus constitutes an invariant.

In general, the weakening rule is applied after every syntax directed
rule.  Applying weakening everywhere does not lead to performance
problems. However, it can be left out in practice at some places to
increase the efficiency of the tool.

The weakening operation $\succeq_{\Gamma}$ is defined in the rule
\textsc{Relax}.  For example, we have
$$
\begin{array}{lll}
P^\text{de} \succeq_{(\cdot)} P^\text{we}
& \equiv &  p^\text{we}_{0,10} \leq p^\text{de}_{0,10} + u_{0,10} - v_{0.10}  \\
& \land &   p^\text{we}_{0} \leq p^\text{de}_{0} - 10 {\cdot} u_{0,10} + 10 {\cdot} v_{0.10}  \\
& \land &  \forall (\alpha, \beta) \neq (0,10)  . \, p^\text{we}_{\alpha,\beta} \leq p^\text{de}_{\alpha,\beta} \;\;\; .
\end{array}
$$

After the constraint generation, the LP solver is provided with a
so-called objective function that it will minimize.  We wish to
minimize the initial potential, which is the resource bound on the
whole program.  Here it is given by $Q$.  Moreover, we would like to
express that minimization of linear potential such as
$$
q_{10,x} |[10,x]|
$$
takes priority over minimization of constant potential such as
$$
q_{0,10} |[0,10]| \; .
$$

To this end, modern LP solvers support efficient resolving of
constraint sets with additional constraints: First we take our initial
constraint set as given in \pref{fig:derivation}.  We then as the
solver to find a solution that satisfies the constraint and minimizes
the linear expression 
$$
q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} \; .
$$
Note the linear penalties (this is a random choice) that we give to
some of the factors.  We use it to express that the interval $[10,x]$
is smaller than the interval $[0,x]$ and, similarly, that the interval
$[x,0]$ is smaller than the interval $[x,10]$.

The LP solver now returns a solution of the constraint set and an
objective value, that is, a mapping from variables to floating-point
numbers and the value of the objective function with this
instantiation.  The solver also memorizes the optimization path that
lead to the optimal solution.  In this case, the objective value would
be
$$
5000
$$
since the LP solver assigns $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.
We now add the constraint
$$
q_{x,0} + 10000{\cdot}q_{0,x} + 10{\cdot}q_{x,10} + 9990q_{10,x} = 5000
$$
to our constraint set and ask the solver to optimize the objective
function
$$
q_0 + 11{\cdot}q_{0,10} \; .
$$
This happens in almost no time in practice.

The final solution is $q_{0,x} = 0.5$ and $q_{*} = 0$ otherwise.  Thus
the derived bound is $0.5 |[0,x]|$.

Another interesting note is that the constraints that we generate have
a particularly simple form that is know as \emph{network problem}.
Such problems can be solve in linear time in practice.
